Is the direction of Granger causal influence the same as the direction of information flow?

### 2015-07-10 paper.tex
Shorten matrix eqn to bring back to 10 pages.
Fix inter-image spacings.
Make \ref -> \eqref as required.

For now, we theoretically evaluate the system parameters. We start with equation~\ref{eq:tx-x-noiseless} and manipulate terms to arrive at an equation bearing the required form of equation~\ref{eq:ar-model-x-x}:
\begin{align*}
	X_i &= \Theta - \widehat\Theta_{i-1} \\
		&= \Theta - \big( \widehat\Theta_{i-2} + \frac{Y_{i-1}}{i-1} \big) \\
		&= \Theta - (\Theta - X_{i-1}) - \frac{X_{i-1} + N_{i-1}}{i-1} \\
		&= X_{i-1} - \frac{X_{i-1} + N_{i-1}}{i-1} \\
		&= \frac{i-2}{i-1} X_{i-1} + \frac{N_{i-1}}{i-1}
\end{align*}
Therefore, $\alpha_1 = \frac{i-2}{i-1}$ and $\epsilon_i = \frac{N_{i-1}}{i-1}$, and hence $\text{Var}(\epsilon_i) = \frac{\sigma_N^2}{(i-1)^2}$.

Next, we model $X_i$ in terms of both the past of $X$ and the past of $\widehat\Theta$:
\begin{equation}
	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \sum_{j=1}^{p}{\beta_j \widehat\Theta_{i-j}} + \tilde\epsilon_i \label{eq:ar-model-x-x-theta}
\end{equation}
We can manipulate equation~\ref{eq:tx-x-noiseless} to bring it into the above form:
\begin{align*}
	X_i &= \Theta - \widehat\Theta_{i-1} \\
		&= X_{i-1} + \widehat\Theta_{i-2} - \widehat\Theta_{i-1}
\end{align*}
Since there is no noise expression here, the Granger causality ratio, $\text{Var}(\epsilon_i) / \text{Var}(\tilde\epsilon_i)$ goes to infinity.

@@

The Granger Causality and Directed Information metrics are a function of the regression coefficients estimated from the data by fitting the models given by equations~\ref{eq:ar-model-x-x} and~\ref{eq:ar-model-x-x-theta}. In our analysis, we described the data-generation model: the algorithm inspired by the Schalkwijk and Kailath scheme for feedback communication. We then proceeded to use the system parameters of this model directly as regression coefficients in our Granger Causality computation. This could be construed as being erroneous: we ought to simulate the data generation, and estimate the regression coefficients from the generated data. This would better model the actions of the neuroscientist who seeks to perform Granger Causality analysis.

@@

Suppose we record two non-stationary processes, $\{X_t\}_{t=1}^n$ and $\{Y_t\}_{t=1}^n$, for which we seek to compute the Granger Causality metrics. To this end, we must find coefficients $\alpha_j(t)$ and $\beta_j(t)$ to minimize the error in fitting the models given by equations~\ref{eq:ar-model-x-x} and~\ref{eq:ar-model-x-x-theta}. Note that $\alpha$ and $\beta$ depend on $t$, since we assume the process is non-stationary. However, since we record the \emph{same} process in each trial, the $\alpha_j(t)$ and $\beta_j(t)$ are constant across trials. Estimating them from data that has many trials is then a simple matter of linear regression.

For a given time instant $t$, the $i$\textsuperscript{th} trial is modeled as
\begin{equation*}
	X_t^{(i)} = \sum_{j=1}^p \alpha_j(t) X_{t-j}^{(i)} + \epsilon_t^{(i)}
\end{equation*}
Note that $\alpha_j(t)$ does not depend on $i$. Collecting variables across $N$ trials, we can write the full model in vector form:
\begin{align*}
	\left[ \begin{array}{c}
			X_t^{(1)} \\
			X_t^{(2)} \\
			\vdots \\
			X_t^{(N)}
	\end{array} \right]
	&=
	\left[ \begin{array}{cccc}
			X_{t-1}^{(1)} & X_{t-2}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			X_{t-1}^{(2)} & X_{t-2}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			\vdots &  & \ddots & \vdots \\
			X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
	\end{array} \right]
	\left[ \begin{array}{c}
			\alpha_1(t) \\
			\alpha_2(t) \\
			\vdots \\
			\alpha_p(t)
	\end{array} \right] \\
	&\qquad \qquad \qquad +
	\left[ \begin{array}{c}
			\epsilon_t^{(1)} \\
			\epsilon_t^{(2)} \\
			\vdots \\
			\epsilon_t^{(N)}
	\end{array} \right]
\end{align*}

If we call the vector on the LHS $\bf{Y}$ and the matrix on the RHS $\bf{X}$, then the vector of $\alpha_j(t)$'s ($\bf{\alpha}(t)$) can be estimated at time instant $t$ using Ordinary Least Squares:
\begin{equation*}
\hat{\bf{\alpha}}(t) = (\bf{X}^T \bf{X})^{-1} \bf{X}^T \bf{Y}
\end{equation*}
This is an unbiased and consistent estimator for $\bf{\alpha}(t)$. This analysis can be trivially extended to the model described by equation~\ref{eq:ar-model-x-x-theta}. With a sufficiently large number of trials, therefore, the system parameters (to be used as regression coefficients in the Granger Causality analysis) can be estimated to arbitrarily high accuracy.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
	\centering
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point20} \\
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point35} \\
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point60}
	\caption{Plots for forward and backward directed information computations for ${\sigma_R^2}/{\sigma_N^2}=0.2$ (top), $0.35$ (center) and $0.6$ (bottom). In each plot, curves for directed information in both directions are illustrated for ratios ${\sigma_\theta^2}/{\sigma_N^2} = 0.25, 0.5,$ and $1$. The x-axis is the number of iterations of message-passing between the transmitter and the receiver. For cases when feedback noise variance $\sigma_R^2$ is moderately smaller than feedforward noise variance $\sigma_N^2$, directed information in the reverse link can dominate that in the forward link. With sufficiently many (albeit sometimes large, as illustrated in the top figure) iterations, directed information in forward information starts to dominate. However, the point at which this happens depends on the (often unknown) ratio of noise in these links. }
		     \label{fig:numerical}
\end{figure}

>>>>

For now, we theoretically evaluate the system parameters. We start with equation~\eqref{eq:tx-x-noiseless} and manipulate terms to arrive at an equation bearing the required form of equation~\eqref{eq:ar-model-x-x}:
\begin{align*}
	X_i &= \Theta - \widehat\Theta_{i-1} \\
		&= \Theta - \big( \widehat\Theta_{i-2} + \frac{Y_{i-1}}{i-1} \big) \\
		&= \Theta - (\Theta - X_{i-1}) - \frac{X_{i-1} + N_{i-1}}{i-1} \\
		&= X_{i-1} - \frac{X_{i-1} + N_{i-1}}{i-1} \\
		&= \frac{i-2}{i-1} X_{i-1} + \frac{N_{i-1}}{i-1}
\end{align*}
Therefore, $\alpha_1 = \frac{i-2}{i-1}$ and $\epsilon_i = \frac{N_{i-1}}{i-1}$, and hence $\text{Var}(\epsilon_i) = \frac{\sigma_N^2}{(i-1)^2}$.

Next, we model $X_i$ in terms of both the past of $X$ and the past of $\widehat\Theta$:
\begin{equation}
	X_i = \sum_{j=1}^{p}{\alpha_j X_{i-j}} + \sum_{j=1}^{p}{\beta_j \widehat\Theta_{i-j}} + \tilde\epsilon_i \label{eq:ar-model-x-x-theta}
\end{equation}
We can manipulate equation~\eqref{eq:tx-x-noiseless} to bring it into the above form:
\begin{align*}
	X_i &= \Theta - \widehat\Theta_{i-1} \\
		&= X_{i-1} + \widehat\Theta_{i-2} - \widehat\Theta_{i-1}
\end{align*}
Since there is no noise expression here, the Granger causality ratio, $\text{Var}(\epsilon_i) / \text{Var}(\tilde\epsilon_i)$ goes to infinity.

@@

The Granger Causality and Directed Information metrics are a function of the regression coefficients estimated from the data by fitting the models given by equations~\eqref{eq:ar-model-x-x} and~\eqref{eq:ar-model-x-x-theta}. In our analysis, we described the data-generation model: the algorithm inspired by the Schalkwijk and Kailath scheme for feedback communication. We then proceeded to use the system parameters of this model directly as regression coefficients in our Granger Causality computation. This could be construed as being erroneous: we ought to simulate the data generation, and estimate the regression coefficients from the generated data. This would better model the actions of the neuroscientist who seeks to perform Granger Causality analysis.

@@

Suppose we record two non-stationary processes, $\{X_t\}_{t=1}^n$ and $\{Y_t\}_{t=1}^n$, for which we seek to compute the Granger Causality metrics. To this end, we must find coefficients $\alpha_j(t)$ and $\beta_j(t)$ to minimize the error in fitting the models given by equations~\eqref{eq:ar-model-x-x} and~\eqref{eq:ar-model-x-x-theta}. Note that $\alpha$ and $\beta$ depend on $t$, since we assume the process is non-stationary. However, since we record the \emph{same} process in each trial, the $\alpha_j(t)$ and $\beta_j(t)$ are constant across trials. Estimating them from data that has many trials is then a simple matter of linear regression.

For a given time instant $t$, the $i$\textsuperscript{th} trial is modeled as
\begin{equation*}
	X_t^{(i)} = \sum_{j=1}^p \alpha_j(t) X_{t-j}^{(i)} + \epsilon_t^{(i)}
\end{equation*}
Note that $\alpha_j(t)$ does not depend on $i$. Collecting variables across $N$ trials, we can write the full model in vector form:
\begin{align*}
	\left[ \begin{array}{c}
			X_t^{(1)} \\
			X_t^{(2)} \\
			\vdots \\
			X_t^{(N)}
	\end{array} \right]
	&=
	\left[ \begin{array}{ccc}
			X_{t-1}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			X_{t-1}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			\vdots & \ddots & \vdots \\
			X_{t-1}^{(N)} & \cdots & X_{t-p}^{(N)} \\
			%X_{t-1}^{(1)} & X_{t-2}^{(1)} & \cdots & X_{t-p}^{(1)} \\
			%X_{t-1}^{(2)} & X_{t-2}^{(2)} & \cdots & X_{t-p}^{(2)} \\
			%\vdots &  & \ddots & \vdots \\
			%X_{t-1}^{(N)} & X_{t-2}^{(N)} & \cdots & X_{t-p}^{(N)} \\
	\end{array} \right]
	\left[ \begin{array}{c}
			\alpha_1(t) \\
			\alpha_2(t) \\
			\vdots \\
			\alpha_p(t)
	\end{array} \right] +
	%&\qquad \qquad \qquad +
	\left[ \begin{array}{c}
			\epsilon_t^{(1)} \\
			\epsilon_t^{(2)} \\
			\vdots \\
			\epsilon_t^{(N)}
	\end{array} \right]
\end{align*}

If we call the vector on the LHS $\bf{Y}$ and the matrix on the RHS $\bf{X}$, then the vector of $\alpha_j(t)$'s ($\bf{\alpha}(t)$) can be estimated at time instant $t$ using Ordinary Least Squares:
\begin{equation*}
\hat{\bf{\alpha}}(t) = (\bf{X}^T \bf{X})^{-1} \bf{X}^T \bf{Y}
\end{equation*}
This is an unbiased and consistent estimator for $\bf{\alpha}(t)$. This analysis can be trivially extended to the model described by equation~\eqref{eq:ar-model-x-x-theta}. With a sufficiently large number of trials, therefore, the system parameters (to be used as regression coefficients in the Granger Causality analysis) can be estimated to arbitrarily high accuracy.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
	\centering
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point20} \\
	\vspace{5mm}
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point35} \\
	\vspace{5mm}
	\includegraphics[width=3.2in]{noisy-fb-sigma_r2-point60}
	\caption{Plots for forward and backward directed information computations for ${\sigma_R^2}/{\sigma_N^2}=0.2$ (top), $0.35$ (center) and $0.6$ (bottom). In each plot, curves for directed information in both directions are illustrated for ratios ${\sigma_\theta^2}/{\sigma_N^2} = 0.25, 0.5,$ and $1$. The x-axis is the number of iterations of message-passing between the transmitter and the receiver. For cases when feedback noise variance $\sigma_R^2$ is moderately smaller than feedforward noise variance $\sigma_N^2$, directed information in the reverse link can dominate that in the forward link. With sufficiently many (albeit sometimes large, as illustrated in the top figure) iterations, directed information in forward information starts to dominate. However, the point at which this happens depends on the (often unknown) ratio of noise in these links. }
		     \label{fig:numerical}
\end{figure}
